Indicators for Ecosystem Condition in Norway

### 2023-10-09 functional_plant_indicators_wetland.Rmd
Print more of the code, add fig captions, and produce scrollable text output

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

## About the underlying data
We use three sets of data for building indicators for ecological condition:

@@

(1) ANO monitoring data:
ANO stands for _areal-representativ naturovervåking_, i.e. _area representative nature monitoring_. 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) NiN reference data:
The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1: < 1/32, 2: 1/32 - 1/8 , 3: 1/8 - 3/8, 4: 3/8 - 4/5, 5: 3/8 - 4/5 + dominance, 6: > 4/5). For the purpose of this project, these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.

(3) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub/tree encroachment, drainage, and pollution.

### Representativity in time and space
For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country. As wetlands occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions. The 1351 plots are distributed across regions in the following way:

@@

### Temporal coverage
The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme. Thus there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.

## Collinearities with other indicators
<!-- Text here -->

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al. 2018). By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for minor ecosystem types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

@@

## Uncertainties
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.

## References
Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

@@

## Analyses {#analyses-fpiw}
### Data sets

ANO data

@@

#### Data handling
- Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)

```{r, include = F}

@@

```{r}
head(ANO.sp.ind)
head(NiN.wetland.cov)
```
For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
For wetland ecosystem we are testing "Light", "Moisture", "Soil_reaction_pH", and "Nitrogen". In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.

@@

```{r, echo = F}
indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {

@@

```{r, echo = T, eval = F}
colnames(NiN.wetland)
# 1st column is the species
# 2nd-46th column is the abundances of sp in different ecosystem types
# 47th-50th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
# each sample re-samples 2/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
                          iter=1000,obl=1,rat=2/3,var.abun=T)

@@

@@

```{r}
head(wetland.ref.cov[[1]])
```

@@

```{r, echo = F}

@@

```{r}
head(wetland.ref.cov.val)
```

@@

We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
```{r, include=FALSE}

@@

```{r}
head(results.wetland[['2-sided']])
```
#### Scaled value analyses

@@

### Ecosystem sub-types
And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
```{r,echo=FALSE,message=FALSE, warning=FALSE}

@@

### Indicator index maps
We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
```{r, echo=FALSE, warning=FALSE,message = FALSE}
#### scaled value maps ####
# keep wide format and add geometry again
res.wetland2 <- results.wetland[['2-sided']]
st_geometry(res.wetland2) <- st_geometry(ANO.wetland)

@@

```
...but here the colors and values of the data points on the map are hard to make out.

>>>>

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE)
```

@@

## About the underlying data {#underlying-data-fpiw}
We use three sets of data for building indicators for ecological condition:

@@

(1) **ANO monitoring data**:
ANO stands for _areal-representativ naturovervåking_, i.e. _area representative nature monitoring_. 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) **NiN reference data**:
The generalized species lists underlying the ecosystem categorization in [NiN](https://artsdatabanken.no/NiN) represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence

    -   1 = \< 1/32

    -   2 = 1/32 - 1/8

    -   3 = 1/8 - 3/8

    -   4 = 3/8 - 4/5

    -   5 = 3/8 - 4/5 + dominance

    -   6 = \> 4/5

For the purpose of this project, these steps are simplified to maximum _expected combination of frequency and cover_, whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.

(3) **The Swedish plant indicator set** published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub/tree encroachment, drainage, and pollution.

### Representativity in time and space {#representativity-fpiw}
For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country. As wetlands occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions. The 1351
<!-- I cannot recreate this number. Use in-line R code, here and elsewhere in similar situations -->
plots are distributed across regions in the following way:

@@

### Temporal coverage {#representativity-temporal-fpiw}
The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme. Thus there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.

## Collinearities with other indicators
Collinearity could occur with other indicators that are responding to the same drivers/pressures, but no really problematic issues have been discussed.

## Reference state and values {#ref-state-fpiw}
### Reference state {#ref-state2-fpiw}
The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al. 2018). By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for minor ecosystem types ( _grunntyper_ or _kartleggingsenheter_ at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

@@

## Uncertainties {#uncertainties-fpiw}
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.

## References {#references-fpiw}
Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

@@

## Analyses {#analyses-fpiw}
### Data sets
<!-- Please print the entire workflow with echo=T. Also, simply reading cached data creates loose ends - I cant find out where this data comes from. You may also chose to read tin this data (or a cached version of it) way at the top of the file, so that you can add some in-line R code for the summary stats in the introduction part (see previous comments)-->
ANO data

@@

#### Data handling
- Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)
<!-- Please include everything like this, but you can chose to have it take up less space by adding the attr.output option, like this: -->

```{r, include = T, attr.output='style="max-height: 300px;"'}

@@

<!-- Print head of data set with horizontal scrolling -->
```{r}
head(ANO.sp.ind) %>%
  kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

```{r}
head(NiN.wetland.cov) %>% 
  kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
For wetland ecosystem we are testing "Light", "Moisture", "Soil_reaction_pH", and "Nitrogen". In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.

@@

```{r, echo = T}
indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {

@@

```{r}
colnames(NiN.wetland)
# 1st column is the species
# 2nd-46th column is the abundances of sp in different ecosystem types
# 47th-50th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
# each sample re-samples 2/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
```

```{r, eval = F}
wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
                          iter=1000,obl=1,rat=2/3,var.abun=T)

@@

<!-- Please insert all the code that produces the cached data. Probably just needs to add the saveRDS() part to the code chunk above. -->

@@

```{r}
head(wetland.ref.cov[[1]]) %>%
    kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

@@

```{r, attr.output='style="max-height: 300px;"'}

@@

```{r}
head(wetland.ref.cov.val) %>%
      kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

@@

We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
```{r}

@@

```{r}
head(results.wetland[['2-sided']]) %>%
      kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```
#### Scaled value analyses

@@

### Ecosystem sub-types {#ecosystem-sub-types-fpiw}
And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
```{r fpiw-plot-maintype,echo=FALSE,message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands. 1 = cases where variable is lower than the reference value- 2 = cases where variablei si higher than the reference value."}

@@

### Indicator index maps
We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...

```{r ph-lower-map-fpiw, echo=FALSE, warning=FALSE,message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands."}
#### scaled value maps ####
# keep wide format and add geometry again
res.wetland2 <- results.wetland[['2-sided']]
st_geometry(res.wetland2) <- st_geometry(ANO.wetland)

@@

...but here the colors and values of the data points on the map are hard to make out.
